Optimal updating factor in adaptive flat-distribution-sampling simulations

### 2017-06-09 doc/invt.tex
intq: logarithmic grid clean up

\item \label{step:qT}
Compute the optimal $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
\footnote{We solved the equation for $\ln q(T)$ using
    a variant of the Newton-Raphson method
    with the bisection method as fallback.
    The initial value set to $q(T) = \ln(1+T\,a_0/2)$.
    Other general methods can be more effective\cite{press3rd}.}
%
%Since $m(Q)$ is a decreasing function of $Q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.\footnote{
  We have used the trapezoidal rule\cite{press3rd}
  on an adaptively defined grid on $[0, q(T)]$ of $N+1$ points
  (we chose $N=20000$)
  such that the resulting change of $\Delta t_{[i]} \equiv t_{[i+1]}-t_{[i]}$
  is no less than $T/(2N)$ and no greater than $2T/N$.
  %
  This requires a few (usually $2$ or $3$) rounds of iterations
  to determine the constant $C$.
  %
  The first guess of $C$ is obtained from integrating
  Eq. \eqref{eq:mint} on a uniform grid.
  %
  Then in each of the next few rounds, we determine the grid points
  as follows.  For grid point $i = 0, 1, \dots, N - 1$,
  we determine the stride
  $\Delta q_{[i]} \equiv q_{[i+1]} - q_{[i]}$
  by a few iterations.
  Starting with the guess, $\bigl(q(T)-q_{[i]}\bigr)/(N-i)$,
  if the resulting increment $\Delta t_{[i]}$
  for grid point $i$ is less than $T/(2N)$,
  we multiply $\Delta q_{[i]}$ by $3/2$
  unless the resulting $q_{[i+1]} > q(T)$,
  or if $\Delta t_{[i]} > 2T/N$,
  we divide $\Delta q_{[i]}$ by $2/3$.
  %
  Once the grid points are defined,
  we will determine the new value of $C$,
  and if the new value differs from the old estimate
  by more than $10\%$, we start another round of iteration.
%
%The curve $q(t)$ is then defined on the grid points,
%$(q_{[i]}, t_{[i]})$,
%
The updating magnitude can then be obtained from numerical differentiation:
    $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$
    and
    $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linear interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.
}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
Mathematically, these updating schemes
allow us to assign zero eigenvalues
to modes that are likely absent from the PMF.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k \le K$,
%
we can construct an updating scheme with
$$
\lambda_0 = \cdots = \lambda_{K-1} = 1,
\mathrm{\; and \;}
\lambda_K = \cdots = \lambda_{n-1} = 0.
$$
%
Then, the updating matrix is given by
%
$$w_{ij} = n \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj},$$
%
and Eq. \eqref{eq:error_asym_singlebin} is modified as
%
\begin{equation}
  \Err_A(T)
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 1 }^{K-1}
    \Gamma_k.
%\notag
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:xt2_eql}.
%
From Eq. \eqref{eq:xt2_eql}, we have
$$E(0) = \frac{1}{2} a_0 \sum_{k=1}^{n-1} \Gamma_k
= \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res1},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 1 }^{ K - 1 }
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This result recovers Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme:
if we set $K + 1 = n$,
%
\begin{equation}
  \sum_{k=1}^{n-1} \Gamma_k
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  ,
  \notag
  %\label{eq:Gamma_G_sum}
\end{equation}
which can be shown by following the steps
similar to those leading to Eq. \eqref{eq:xt2_sum}.
%
%
%
\note{The derivation.
From Eq. \eqref{eq:eig_orthonormal_cols}, we have
$\sum_{i=1}^n \rho_i \langle v_{*i}(t) \, v_{*i}(0) \rangle
=\sum_{k=1}^{n-1} \langle {\tilde v}_{*k}(t) \, {\tilde v}_{*k}(0) \rangle$,
so
\begin{align*}
\sum_{i=1}^n \rho_i G_i
&=\sum_{t=-\infty}^\infty\sum_{i=1}^n
  \rho_i \langle v_{*i}(t) \, v_{*i}(0) \rangle \\
  &=\sum_{t=-\infty}^\infty\sum_{k=1}^{n-1}
  \langle {\tilde v}_{*k}(t) \, {\tilde v}_{*k}(0) \rangle
  =\sum_{k=1}^{n-1} \Gamma_k.
\end{align*}
}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.

>>>>

\item \label{step:qT}
Compute the optimal $q(T)$ by solving Eq. \eqref{eq:opt_qT}.
\footnote{We solved the equation for $\ln q(T)$ using
    a variant of the Newton-Raphson method
    with the bisection method as a fallback.
    The initial value is set to $q(T) = \ln(1+T\,a_0/2)$.}
%
%Since $m(Q)$ is a decreasing function of $Q$,
%we can start from an interval $[Q_L, Q_R]$
%that satisfies $m(Q_L) > 2/(T\,a_0) > m(Q_R)$
%(e.g. we can choose $Q_L = 0$,
%  and repeatedly double $Q_R$ until
%  the latter inequality is satisfied).
%%
%We can then iteratively refine the interval
%by substituting $Q_M = (Q_L + Q_R)/2$
%for either $Q_L$ and $Q_R$,
%depending on the sign of $m(Q_M) - 2/(T \, a_0)$,
%until the width of the interval is less than
%the desired precision,
%%
%and set $q(T)$ to $Q_M$.

\item \label{step:alpha}
Compute the optimal schedule by
  numerically integrating Eqs. \eqref{eq:q_opt} and \eqref{eq:mint}.\footnote{
  We have used the trapezoidal rule
  on a $N+1$ point logarithm grid over $[0, q(T)]$:
    $q_{[i]} = \bigl(q(T) + 1\bigr) - \bigl(q(T)+1\bigr)^{1-i/N}$,
  where $i = 0, \dots, N$, and we chose $N=20000$.
%
The integration yields a curve $q(t)$ on the grid points,
$(q_{[i]}, t_{[i]})$.
%
The updating magnitude can then be obtained from numerical differentiation:
    $\alpha_i = (1 - e^{-\lambda_1 \, \hat \alpha_i})/\lambda_1$
    and
    $\hat \alpha_i \equiv (q_{[i+1]} - q_{[i-1]}) / (t_{[i+1]} - t_{[i-1]})$,
with
$q_{[ -1]} \equiv 0$,
$q_{[N+1]} \equiv q(T)$,
$t_{[ -1]} \equiv 0$, and
$t_{[N+1]} \equiv T$.
%
The resulting values of $(t_{[i]}, \alpha_{[i]})$
can be linear interpolated
to get the updating magnitude, $\alpha(t)$, at any simulation time, $t$.
}

@@

We can generalize
the single-bin scheme to a class of
perfect (brick-wall) \emph{bandpass} updating schemes
that ignore short-wavelength eigenmodes,
which are typically local noise.
%
Mathematically, these updating schemes
allow us to assign zero eigenvalues
to modes that are likely absent from the PMF.
%
For example,
%if we can assume
by assuming
that the PMF is sufficiently smooth
such that the eigenmodes can be limited to $k < K$,
%
we can construct an updating scheme with
$$
\lambda_0 = \cdots = \lambda_{K-1} = 1,
\mathrm{\; and \;}
\lambda_K = \cdots = \lambda_{n-1} = 0.
$$
%
Then, the updating matrix is given by
%
$$w_{ij} = n \sum_{k=0}^{K-1} \phi_{ki} \, \phi_{kj},$$
%
and Eq. \eqref{eq:error_asym_singlebin} is modified as
%
\begin{equation}
  \Err_A(T)
  =
  \frac {       T     }
        { (T + t_0)^2 }
  \sum_{ k = 1 }^{K-1}
    \Gamma_k.
%\notag
\label{eq:error_asym_sinc}
\end{equation}
%
The optimal schedule of these schemes
remains to be the inverse-time schedule
given by Eq. \eqref{eq:alpha_invt1},
%
with
%
\begin{equation}
  t_0
  =
  \frac{        1      }
       { \alpha(t = 0) }
  =
  \frac{  2  }
       { a_0 }
  ,
\label{eq:t0_sinc}
\end{equation}
%
where we have used Eq. \eqref{eq:half_alpha0}
in the second step.
%
\note{
For the single-bin updating scheme,
it recovers Eq. \eqref{eq:t0_singlebin}
under the assumption of Eq. \eqref{eq:xt2_eql}.
%
From Eq. \eqref{eq:xt2_eql}, we have
$$E(0) = \frac{1}{2} a_0 \sum_{k=1}^{n-1} \Gamma_k
= \frac{1}{2} \, a_0 \, \Gamma.$$
}
%
From Eqs. \eqref{eq:error_tot},
\eqref{eq:error_res},
\eqref{eq:tk0},
and
\eqref{eq:error_asym_sinc},
we get
%
\begin{equation}
  \Err(T)
  =
  \frac{   1     }
       { T + t_0 }
  \sum_{ k = 1 }^{ K - 1 }
    \Gamma_k
  .
\label{eq:error_sinc}
\end{equation}
%
This result recovers Eq. \eqref{eq:Emin_singlebin}
for the single-bin updating scheme:
if we set $K = n$, then\note{
  From Eq. \eqref{eq:eig_orthonormal_cols}, we have
  $\sum_{i=1}^n \rho_i \langle v_{*i}(t) \, v_{*i}(0) \rangle =\sum_{k=1}^{n-1} \langle {\tilde v}_k(t) \, {\tilde v}_k(0) \rangle.$
  Then sum over $t$ from $\infty$ to $\infty$.
}
%
\begin{equation}
  \sum_{k=1}^{n-1} \Gamma_k
  = \sum_{i=1}^n \rho_i \, G_i = \Gamma
  ,
  \notag
  %\label{eq:Gamma_G_sum}
\end{equation}
%
This error is roughly proportional to the inverse simulation length,
as one would expect from an equilibrium FDS simulation.
%In Appendix \ref{sec:equilerr},
%we show that the above error tends
%to the error of the ideal equilibrium FDS simulation
%in the asymptotic limit.
